---
title: "UMAP with tidymodels"
author: "Pietro Franceschi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)

```


## Introduction

The objective of this demo is to demonstrate how how to use UMAP by using `tidymodels`. 

The steps needed to include UMAP projection in a recipe are included in the `embed` package with also include other methods for embedding predictors in new columns. Also *sparse pca* is included there

```{r}
library(embed)
```




Here we get the data
```{r}
load("data/KOMP_data_targeted.RData")
komp
```

And now imputation!

```{r}
komp_recipe <- recipe(~., data = komp) %>% 
  update_role(c("MouseID","Genotype","Gender","Zygosity"), new_role = "id") %>% 
  step_mutate_at(all_numeric_predictors(), 
                 fn = ~ ifelse(is.na(.), min(., na.rm = TRUE), .)) %>% 
  step_impute_lower(all_predictors()) %>%
  step_log(all_predictors()) %>% 
  #step_normalize(all_predictors()) %>%  
  step_umap(all_predictors(), neighbors = 10) %>% 
  prep()
```


Let's give a look to the matrix after imputation ...

And now let's display the results


```{r}
komp_recipe %>% 
  juice() %>% 
  ggplot() + 
  geom_point(aes(x = UMAP1, y = UMAP2, col = Gender), size = 2, alpha = 0.7) + 
  scale_color_brewer(palette = "Set1") + 
  ggtitle("Komp UMAP") + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

We see many groups. Not clear unfortunately. Now it would be necessary to discuss the outcomes with the chemical analyst, the biochemists ...


## Something for you

* Go through the code and understand what I did
* Try different methods of scaling and imputing (eg. knn `step_impute_knn`)
* Try UMAP on the MTBLS1 dataset


```{r}
library(readr)
MTBSL1 <- read_delim("data/MTBSL1.tsv", delim = "\t", 
    escape_double = FALSE, trim_ws = TRUE)
```



```{r}
MTBLS1_recipe <- recipe(~., data = MTBSL1) %>% 
  update_role(!all_numeric(), new_role = "id") %>% 
  step_zv(all_predictors()) %>% 
  step_center(all_predictors()) %>% 
  step_pca(all_predictors()) %>% 
  prep()
  
```


```{r}
MTBLS1_recipe_umap <- recipe(~., data = MTBSL1) %>% 
  update_role(!all_numeric(), new_role = "id") %>% 
  step_zv(all_predictors()) %>% 
  step_center(all_predictors()) %>% 
  step_umap(all_predictors(), neighbors = 7) %>% 
  prep()
  
```





```{r}
MTBLS1_recipe_umap %>% 
  juice() %>% 
  ggplot() + 
  geom_point(aes(x = UMAP1, y = UMAP2, col = Gender, shape = Metabolic_syndrome), size = 2, alpha = 0.7) + 
  scale_color_brewer(palette = "Set1") + 
  ggtitle("Iris Scoreplot") + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

